Sequential rank agreement methods for comparison of ranked lists

### 2016-04-19 latex/superranker-manuscript.tex
new version of section 2 in new file superranker-manuscript.tex

\begin{document}
\input{paper-concordance}

@@

@@

Consider a set of $P$ different items $X=\{X_1,\dots,X_P\}$ and a
ranking function $R: \{X_1,\dots,X_P\}\to \{1,\dots,P\}$, such that
$R(X_p)$ is the rank of item $X_p$. The inverse mapping $R^{-1}$ gives
the item $R^{-1}(r)$ that was assigned to rank $r\in\{1,\dots,P\}$. An
ordered list is the realization of a ranking function $R$ applied to a
data sample $X$ and we let $R_1(X),\dots,R_L(X)$, $L\geq2$, denote
ordered lists obtained from $L$ different ranking functions applied to
the same data $X$. We will use the same notation for the equivalent
scenario where we have ordered lists obtained from applying the same
ranking function to $L$ samples from the same population, i.e., for
ranked lists $R(X^1), \ldots, R(X^L)$.  Panels (a) and (b) of
Table~\ref{tab:example} show a schematic example of these
mappings. Thus if $R_l^{-1}(1)=X_{34}$ then item $X_{34}$ is ranked
first in list $l$ and similarly $R_l(X_{34})=1$.

@@

The agreement of the lists regarding the rank given to an item $X_p$ can be
measured by
\begin{equation}
  A(X_p) = f(R_1(X_p), \ldots, R_L(X_p))
\end{equation}
for a measurable non-negative real-valued distance function $f$.
For an integer $1\le d\le P$ we define the unique set of items found
by merging the first $d$ elements of each of the $L$ lists, i.e., the
set of items ranked less than or equal to $d$ in any of the lists:
\begin{equation}
S_d = \{R_l^{-1}(r) ; r \leq d, l = 1, \ldots, L \},
\end{equation}
which is exemplified in Panel (c) of Table \ref{tab:example}.
We then define the \emph{sequential rank agreement} as the weighted expected agreement of the items
found in the set $S_d$:
\begin{align}
\textrm{sra}(d) &= \sum_{\{p \in S_d\}}|S_d|^{-1} \E[A(X_p)]
\end{align}
where $|S_d|$ refers to the cardinality of the set $S_d$, and the expectation
is taken with respect to the distribution of the rankings of $X_p$ accross the $L$ lists.
Values of \textrm{sra} close to zero suggests that the lists agree while larger
values suggest disagreement. Just like in method comparison studies we
prefer smaller limits of agreement because that means that the
differences between the methods is small. If the ranked lists are
identical then the value of sequential rank agreement will be zero for
all depths $d$.

%Note that the terms that are part of \eqref{def:sra} are not
%independent since every list contains the ranks from 1 to $P$ exactly
%once. As a consequence we simply use the pooled standard deviation as
%a measure of the average distance between rankings and we refrain from
%using traditional distributional results for this measure.

Throughout the rest of this paper we will use the variance to quantify the
agreement of the rankings and therefore define
\begin{align}
  A(X_p) = (R_i(X_p) -  \E[R_i(X_p)])^2
\end{align}
For each item, the function $A$ has an interpretation as the expected
euclidean distance of the individual rankings from the expected ranking over all the lists, and
it corresponds to the same measure that is used in method comparison studies to compute the
limits of agreement \citep{alt:bland:1983}. The sequential rank agreement then becomes equivalent
to the pooled variance of the items found in $S_d$, and it can be interpreted as
the average distance of the individual rankings of the lists from the
average ranking for each of the items we have seen until depth $d$.

Given a sample of $1,\ldots,L$ lists an estimator of the sequatial rank agreement is given by
\begin{equation}
\widehat{\textrm{sra}}_L(d)= \frac{\sum_{\{p \in S_d\}}(L-1)\widehat{A_L}(X_p)}{(L-1)|S_d|}\label{def:sra}
\end{equation}
where
\begin{align}
\widehat{A}_L(X_p) = \frac{1}{L-1}\sum_{i=1}^L (R_i(X_p) - \overline{R}(X_p))^2, \quad \overline{R}_L(X_p) = \frac{1}{L}\sum_{i=1}^L R_i(X_p)
\end{align}
is the sample variance of the rank of item $X_p$. For an easier interpretation we suggest to plot $\widehat{\textrm{sra}}_L(d)^{1/2}$ corresponding to the pooled standard deviation. The consistency of $\widehat{\textrm{sra}}_L$ as $L \rightarrow \infty$ will be discussed further in section \ref{sec:chgpoint}.

@@

% latex table generated in R 3.2.3 by xtable 1.8-2 package
% Fri Apr  8 10:22:58 2016
\begin{table}[tb]
\centering
\caption{Top 10 list of ranked results from the Golub data. Numbers indicate the predictor/gene for the given ranking and method} 
\label{tab1}
\begin{tabular}{rrrrr}
  \hline
Ranking & T & LogReg & ElasticNet & MIC \\ 
  \hline
1 & 2124 & 2124 & 829 & 378 \\ 
  2 & 896 & 896 & 2124 & 829 \\ 
  3 & 2600 & 829 & 2198 & 896 \\ 
  4 & 766 & 394 & 1665 & 1037 \\ 
  5 & 829 & 766 & 1920 & 2124 \\ 
  6 & 2851 & 2670 & 1042 & 808 \\ 
  7 & 703 & 2939 & 808 & 108 \\ 
  8 & 2386 & 2386 & 849 & 515 \\ 
  9 & 2645 & 1834 & 937 & 2670 \\ 
  10 & 2002 & 378 & 1995 & 2600 \\ 
   \hline
\end{tabular}
\end{table}

@@

>>>>

\begin{document}

@@

% chunk 1

@@

Consider a set of $P$ different items $X=\{X_1,\dots,X_P\}$ and a
ranking function $R: \{X_1,\dots,X_P\}\to \{1,\dots,P\}$, such that
$R(X_p)$ is the rank of item $X_p$. The inverse mapping $R^{-1}$ gives
the item $R^{-1}(r)$ that was assigned to rank $r\in\{1,\dots,P\}$. An
ordered list is the realization of a ranking function $R$ applied to
the set of items $X$. In all what follows we condition on the set of
items and consider the ranking function to be a random variable. Thus,
let $R_1(X),\dots,R_L(X)$, $L\geq2$, be a sample of $L$ independent
identically distributed draws from an unknown probability distribution
function $Q$. One aim is then to test how much $Q$ resembles the
uniform distribution which assigns probability $(1/P)^P$ to each of
the $P^P$ different rankings.

\bigskip\hrule HOV: er det virkelig equivalent ogsaa mht vores nye theoretiske
ideer? Og bruger vi det nogen steder?  enten skal det forklares med et
konkret eksempel eller ud?  We will use the same notation for the
equivalent scenario where we have ordered lists obtained from applying
the same ranking function to $L$ samples from the same population,
i.e., for ranked lists $R(X^1), \ldots, R(X^L)$.  
\hrule\bigskip

Panels (a) and (b) of
Table~\ref{tab:example} show a schematic example of these
mappings. Thus if $R_l^{-1}(1)=X_{34}$ then item $X_{34}$ is ranked
first in list $l$ and similarly $R_l(X_{34})=1$.

@@

The agreement of the lists regarding the rank given to an item $X_p$
can be measured by the expected variance across the lists:
\begin{equation}
  {A}(X_p) = \sum_{r=1}^{P^P} \{r(X_P)-\E_Q R(X_P)\}^2 Q(r)
\end{equation}
where $\E_Q R(X_p)=\sum_{r=1}^{P^P} r(X_p)Q(r)$. The empirical
counterpart is

\begin{align}
\widehat{A}_L(X_p) = \frac{1}{L-1}\sum_{i=1}^L (R_i(X_p) - \overline{R}(X_p))^2, \quad \overline{R}_L(X_p) = \frac{1}{L}\sum_{i=1}^L R_i(X_p).
\end{align}
For each item, the function $\widehat{A}_L$ has an interpretation as the expected
Euclidean distance of the individual rankings from the expected
ranking over the $L$ lists, and it corresponds to the same measure
that is used in method comparison studies to compute the limits of
agreement \citep{alt:bland:1983}.

\bigskip\hrule HOV
Vi kunne sige her: Alternative distance functions (f non-negativ real valued) are discussed in section ??
\hrule\bigskip 
% \begin{equation}
  % \widehat{A_L}(X_p) = f(R_1(X_p), \ldots, R_L(X_p))
% \end{equation}
% for a measurable non-negative real-valued distance function $f$.
For an integer $1\le d\le P$ we define the expected set of unique
items found by merging the first $d$ elements across the possible
lists:
\begin{equation}
S_d = \{X_p; \left(\sum_{r=1}^{P^P} I\{r(X_p)\le d\} Q(r)\right) >0\}.
\end{equation}
The empirical counterpart the set of unique items ranked less than or
equal to $d$ in any of the $L$ lists:
\begin{equation}
\widehat{S}_{d,L} = \{X_p; \left(\sum_{l=1}^L I\{R_l(X_p)\le d\}\right)>0\},
\end{equation}
which is exemplified in Panel (c) of Table \ref{tab:example}.

We then define the \emph{sequential rank agreement} as the weighted
expected agreement of the items found in the set $S_d$:
\begin{equation}
  \textrm{sra}(d) = \frac{\sum_{\{p \in S_d\}}A(X_p)}{|S_d|}\label{def:sra}
  % & =\sum_{p=1}^P \sum_{l=1}^L I\{R_l(X_p)\le d\}>0\}\notag
\end{equation}
 where $|S_d|$ is the cardinality of the set $S_d$.
The empirical counterpart is given by
\begin{equation}
  \widehat{\textrm{sra}}_L(d) = \frac{\sum_{\{p \in \widehat{S}_{d,L}\}}(L-1)\widehat{A_L}(X_p)}{(L-1)|\widehat {S}_{d,L}|}.\label{def:sra}
                                % \intertext{where}
                                % |S_d| & =\sum_{p=1}^P \sum_{l=1}^L I\{R_l(X_p)\le d\}>0\}\notag
\end{equation}
% where is the cardinality of the set $\widehat{S}_{d,L}$.

Values of \textrm{sra} close to zero suggests that the lists agree
while larger values suggest disagreement.

Just like in method comparison studies we prefer smaller limits of
agreement because that means that the differences between the methods
is small. If the ranked lists are identical then the value of
sequential rank agreement will be zero for all depths $d$.
%Note that the terms that are part of \eqref{def:sra} are not
%independent since every list contains the ranks from 1 to $P$ exactly
%once. As a consequence we simply use the pooled standard deviation as
%a measure of the average distance between rankings and we refrain from
%using traditional distributional results for this measure.

The sequential rank agreement then becomes equivalent to the pooled
variance of the items found in $S_d$, and it can be interpreted as the
average distance of the individual rankings of the lists from the
average ranking for each of the items we have seen until depth $d$.

\subsection{Population parameter}

Given a sample of $1,\ldots,L$ lists an estimator of the sequential
rank agreement is given by
\begin{equation}
\widehat{\textrm{sra}}_L(d)= \frac{\sum_{\{p \in S_d\}}(L-1)\widehat{A_L}(X_p)}{(L-1)|S_d|}\label{def:sra}
\end{equation}
where
\begin{align}
\widehat{A}_L(X_p) = \frac{1}{L-1}\sum_{i=1}^L (R_i(X_p) - \overline{R}(X_p))^2, \quad \overline{R}_L(X_p) = \frac{1}{L}\sum_{i=1}^L R_i(X_p)
\end{align}
is the sample variance of the rank of item $X_p$. For an easier interpretation we suggest to plot $\widehat{\textrm{sra}}_L(d)^{1/2}$ corresponding to the pooled standard deviation. The consistency of $\widehat{\textrm{sra}}_L$ as $L \rightarrow \infty$ will be discussed further in section \ref{sec:chgpoint}.

@@

% chunks 2,3,4,5,6,7

@@

% chunks 8,9
